---
title: Baby Names
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
```

## BabyNames

### Get Data

```{r}
bnames = read_csv('../../data/babynames.csv.gz')
bnames
```

## Top Female Names

1. [x] Filter the data for sex = "F"
       Filter the data for year: year >= 2010 and year <= 2020
2. [x] Group By "name"
3. [x] For each name, sum all values of n.
4. [x] Arrange the names in descending order of births.
5. [x] Extract the top 10 names.
6. [x] Make a barplot of the top names.

```{r}
bnames %>% 
  filter(sex == "F") %>% 
  group_by(name) %>% 
  summarize(
    nb_births = sum(n)
  ) %>% 
  arrange(desc(nb_births)) %>% 
  head(10) %>% 
  mutate(name = forcats::fct_reorder(name, nb_births)) %>% 
  ggplot(aes(y = name, x = nb_births)) +
  geom_col()
```


```{r}
bnames %>% 
  filter(sex == "F") %>% 
  group_by(name) %>% 
  summarize(
    nb_births = sum(n)
  ) %>% 
  arrange(desc(nb_births)) %>% 
  head(10) %>% 
  mutate(name = forcats::fct_reorder(name, nb_births)) %>% 
  ggplot(aes(y = name, x = nb_births)) +
  geom_col()
```


1. [x] Filter the data for sex = "F"
       Filter the data for year: year >= 1980
2. [x] Add a new variable called decade
2. [x] Group By "decade" and "name"
3. [x] For each name, sum all values of n.
4. [x] Arrange the names in descending order of births.
5. [x] Extract the top 10 names for each decade.
6. [ ] Make a barplot of the top names by decade

```{r}
# 2014 -> 2010
round(2014/10)*10
```


```{r}
bnames %>% 
  filter(year > 1980) %>% 
  filter(sex == "F") %>% 
  mutate(decade = round(year/10)*10) %>% 
  group_by(decade, name) %>% 
  summarize(
    nb_births = sum(n)
  ) %>% 
  arrange(desc(nb_births)) %>% 
  top_n(10, nb_births) %>% 
  mutate(name = forcats::fct_reorder(name, nb_births)) %>% 
  ggplot(aes(y = name, x = nb_births)) +
  geom_col() +
  facet_wrap(~ decade, scales = 'free_y')
```

## Shiny

```{r}
library(shiny)
options("shiny.launch.browser" = rstudioapi::viewer)
ui <- fluidPage(
  # [x] Add a text input to let the user enter their name
  textInput('name', 'Enter name'),
  # Add a text output to display a greeting
  textOutput('greeting')
)

server <- function(input, output, session){
  # Render the greeting
  output$greeting <- renderText({
    paste("Hello, ", input$name)
  })
}

shinyApp(ui, server)
```

```{r}
output$plot_10_names = 
```


## App

1. Input: User selects
   - starting year (1980)
   - sex ('M' or 'F')
2. Output:
   - top 10 names
   
```{r}
ui <- fluidPage(
  # Add an input to select starting year
  sliderInput('year_start', 'Enter starting year', value = 1900, 
      min = 1900, max = 2019, animate = TRUE, step = 10
  ),
  # Add an input to select sex
  selectInput('sex', 'Select sex', choices = c('M', 'F')),
  # Add an output to display a plot
  plotOutput('plot_top_10_names')
)

server <- function(input, output, session){
  output$plot_top_10_names = renderPlot({
    year_selected = input$year_start
    sex_selected = input$sex
    bnames %>% 
      filter(year > year_selected) %>% 
      filter(sex == sex_selected) %>% 
      group_by(name) %>% 
      summarize(
        nb_births = sum(n)
      ) %>% 
      arrange(desc(nb_births)) %>% 
      head(10) %>% 
      mutate(name = forcats::fct_reorder(name, nb_births)) %>% 
      ggplot(aes(y = name, x = nb_births)) +
      geom_col()
  })
}

shinyApp(ui, server)
```



1. Separate by 10 year periods and most common name?
2. Average length of a name by year?
3. How many babies have name starting with 'S'?
4. Most common unisex names.
4. What will be the most popular names 10 years in the future?